Pollution, agricultural productivity, and development: Evidence from coal in plants in India\footnote{All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}

### 2023-07-13 draft/draft.tex
new draft

\title{Air pollution and agricultural productivity in a developing country\footnote{I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}}
\author{Joshua D. Merfeld\footnote{KDI School of Public Policy and Management and IZA; \href{mailto:merfeld@kdis.ac.kr}{\nolinkurl{merfeld@kdis.ac.kr}}}}
\date{2023-07-10}

@@

While there are likely positive effects from this increase in power generation -- for example, electricity may lead to economic and productivity growth (Dinkelman 2011; Kline and Moretti 2014; Rud 2012; Van de Walle et al. 2017), though electrification alone may not be a sufficient condition (K. Lee, Miguel, and Wolfram 2020) -- these improvements do not come without downsides, especially when it comes to power derived from coal. Pollutants from coal contribute to climate change and the phase out of coal energy is therefore a major climate goal (IEA 2022). Emissions from the production and burning of coal are also directly harmful to human health and the environment; pollutants like sulfur dioxide, nitrogen oxides, methane, and mercury are particularly harmful.\footnote{www.eia.gov/energyexplained/coal/coal-and-the-environment.php} While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.\footnote{www.npr.org/2022/08/15/1117560560/a-rising-demand-for-coal-amidst-war-in-ukraine}

@@

In this paper, I study the effects of pollution on agricultural productivity in India. Previous work has suggested that there may be harmful direct effects of air pollution (Heck et al. 1982; Marshall et al. 1997) and water pollution (Reddy and Behera 2006) on agricultural productivity, including work showing that gold mines -- which release large amounts of pollutants into the surrounding environment -- lead to lower agricultural productivity in Ghana (Aragón and Rud 2016). These effects can work through the well-known effects on human health and productivity, but also directly through land productivity. Black carbon in soot, for example, can directly affect solar radiation reaching plants on the surface (Ramanathan and Carmichael 2008; Burney and Ramanathan 2014). Relative to previous work in economics, this paper tackles pollution more broadly, before turning to a specific source of pollution: coal plants. I show that air pollution leads to lower agricultural productivity in India and that pollution from coal plants is particularly harmful to agriculture.

@@

The main goal of this paper is to examine whether exposure to pollution and emissions from coal plants, in particular, affect agricultural productivity. I use several sources of data, which I describe here and list in Table \ref{tab:data} of the appendix.

@@

The second dataset includes agricultural productivity for both the monsoon and winter seasons, from 2002 to 2018. To match other data, I only use the data up to 2013. This data comes from Gangopadhyay et al. (2022) and estimates land productivity (i.e.~yield, in tons per hectare) for the major crops in India, using satellite-derived data. While there is a wealth of literature on measurement error in land (Desiere and Jolliffe 2018; Abay et al. 2022), the authors use a static cropland mask for the productivity estimates. This means that any measurement error will be constant across time and subsumed by the village fixed effects. While the measure of output is also noisy, the use of satellite-derived data and a static cropland mask reduces concerns about measurement error being responsible for the results below, especially since common sources of measurement error are unlikely to be correlated with wind direction. The authors define the monsoon season as June to October and the winter season as November to March. I focus on only the monsoon season and keep the authors' definition when matching across data below. This data is also publicly available from Nature's data-sharing website.\footnote{springernature.figshare.com} Since the main regressions below use village fixed effects, any differences in cropping patterns across areas should not bias estimates on changes, as long as these patterns do not change markedly in response to pollution.

@@

To measure exposure to pollution from coal plants, I first locate all village centroids located within 30km from a coal plant in a given year. I choose 30km due to previous research on the effects of (air) pollution (Aragón and Rud 2016; Bencko and Symon 1977; Li and Gibson 2014). I then calculate the direction from coal plants to all village centroids within that 30km radius. To define exposure, I then pull daily wind direction data from the National Center for Atmospheric Research.\footnote{climatedataguide.ucar.edu/} For each day, I document whether the wind is blowing towards each village centroid\footnote{I define ``towards'' as within five degrees to help take into account that I am using village centroids, using the x and y components of wind speed to calculate angles.} and then temporally aggregate this daily data depending on the temporal definition of the corresponding outcome. For example, agricultural productivity is defined across five months (e.g.~the monsoon season is from June to October) so I count the total days a given village is exposed to wind blowing from \emph{any} coal plant within 30km during those five months.

The top panel in Figure \ref{fig:windexample} in the appendix is an example wind direction raster. The raster shows wind direction for the entirety of India on January 1st, 2010, as well as the location of coal power plants on that date. The prevailing winds on the date differ across the country and, though not shown in the figure, across days. This means that the overall exposure for a given area to coal plant emissions changes across time. I also pull data on particulate matter to help validate this proxy for exposure. This data comes from Hammer et al. (2020) and I also aggregate this to the village using the same method detailed above.

To help understand some of the mechanisms driving the relationship between exposure to emissions and agricultural productivity, I also look at the differential effects by the labor intensity of the predominant crop in each area. For this, I use data from (Monfreda, Ramankutty, and Foley 2008), who estimate area harvested for different crops across the global around the year 2000. I identify areas where rice (paddy) is the most commonly grown crop, under the assumption that rice is more labor intensive than other crops (Vollrath 2011; Michler 2020). Michler (2020), for example, shows that rice, on avearge, uses almost 1,800 hours per hectare, while the second-highest crop in his data, cotton, juses just 850.\footnote{He estimates that wheat, another common crop in India, uses just 600 hours per hectare.}

@@

\caption{\label{tab:plantresultstable}Local characteristics and the construction of coal plants}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{1991 census} & \multicolumn{2}{c}{2001 census} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
  & 1991 & 2001 & 2001 & 2011\\
\midrule
Population (log) & 0.012*** & 0.009*** & 0.015*** & 0.006***\\
 & (0.001) & (0.000) & (0.000) & \vphantom{1} (0.000)\\
Area (log) & -0.010*** & -0.005*** & -0.010*** & -0.005***\\
 & (0.001) & (0.000) & (0.000) & (0.000)\\
Urban pop (prop) & -0.029 & 0.006 & -0.003 & -0.007\\
 & (0.037) & (0.035) & (0.022) & (0.018)\\
Ag. productivity &  &  & 0.014*** & 0.028***\\
 &  &  & (0.001) & (0.001)\\
Pollution &  &  & 0.051*** & 0.016***\\
(log PM 2.5) &  &  & (0.002) & (0.001)\\
Sub-sample & all & no plant & all & no plant\\
\midrule
Observations & 283,021 & 264,265 & 508,786 & 469,005\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: State fixed effects are included in regressions. Robust standard errors are in parentheses. The outcome in the first column is whether the village is within 30km of a coal plant in 1991. The second column is whether a village in 1991 will have a coal plant in 2001, conditional on not having one in 1991. The last two columns are similarly defined, except using 2001 and 2011 as the years. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

The inclusion of rainfall and temperature is meant to reduce concerns that wind direction may not affect agricultural productivity but, instead, be correlated with it through changes in weather. For example, prevailing winds from one direction may be associated with dry weather, while winds from another direction may be more likely to carry moisture and rain. It is not a priori clear how this would bias the estimates with respect to pollution and coal plants, however. The differences in weather patterns is very context specific and can differ even for villages located near one another. Nonetheless, adding rainfall and temperature should alleviate some of these concerns.

It is worth taking the time to validate the use of wind direction as a measure of exposure to pollution. Consider the data used for the regressions presented in Table \ref{tab:yield3ivmainfirststage} of the appendix. The outcome variable is particulate matter -- specifically, \(\mathrm{PM 2.5}\), which is particulate matter no larger than 2.5 micrometers in diameter -- which comes from Hammer et al. (2020). \(\mathrm{PM 2.5}\) is one of the harmful byproducts from coal plants, along with sulfur dioxide, different types of nitrogen oxides, and mercury.\footnote{www.epa.gov/airmarkets/power-plants-and-neighboring-communities} This data is available at the monthly level, but I aggregate variables to the season variable since that is the level of analysis in the main results.

Both the particulate matter data and the wind data is at the village level, which means the regressions in Table \ref{tab:yield3ivmainfirststage} are at the village level. As such, the geographic fixed effects are village fixed effects and the temporal fixed effects are month fixed effects. Since the exposure variable is at the village level and we follow villages over time, the standard errors are also clustered at the village level. In columns two and four, I switch the month fixed effects for district-month fixed effects, which considers only villages in the same district in the same month.

In addition to serving as an example, Table \ref{tab:yield3ivmainfirststage} also helps validate the use of wind direction as a proxy for exposure to pollution. The first two columns include all available years of data while the last two columns restrict estimation only to the years for which the agricultural productivity regressions are estimated. Across all four columns, the story is the same: when the wind is blowing in the direction of a village from a coal plant within 30 kilometers, estimated \(\mathrm{PM 2.5}\) is substantially higher.

@@

To identify possible mechanisms, I identify areas that grow predominantly rice, which is a labor-intensive crop. Specifically, we are interested in the following regression:
\begin{gather} 
y_{it} = \alpha_{i} + \gamma_{t} + \omega PM_{it} + \phi rice_{it} + \beta PM_{it}\times rice_{it} + \delta rain_{it} + \psi temp_{it} + \varepsilon_{it},
\end{gather}
where \(PM_{it}\) denotes the effect of pollution in non-rice areas and \(PM_{it}\times rice_{it}\) is the differential effect in rice-growing areas. Here, I create an indicator variable for rice areas that is time invariant, so \(rice_{it}\) drops out of the estimation. I instrument for \(PM_{it}\) and for \(PM_{it}\times rice_{it}\) using wind and wind interacted with the indicator for rice.

@@

I first present the reduced form results for agricultural productivity. The outcome of all regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. The first set of results are in Table \ref{tab:yieldtable}. The unit of analysis is the village-season -- from 2002 to 2013 -- focusing only on the monsoon (kharif) season. The data used to construct the yield measure is calculated using five months of data, so the wind variable is defined for the same five months and is the sum of days in which the wind is pointed towards the village centroid from a coal plant.

The first column presents the most simple results, with only the wind exposure variable and the fixed effects. The second column adds a rainfall and average temperature variable, while the third column includes higher-order weather variables, as well. While the coefficient decreases slightly in magnitude as we move from the first to the second column and the second to the third column, the decrease is relatively small. There is a consistent, negative effect of wind exposure on agricultural productivity.

@@

\caption{\label{tab:yieldtable}Wind direction and agricultural productivity}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
\toprule
  & (1) & (2) & (3)\\
\midrule
Wind days & -0.0078*** & -0.0065*** & -0.0051***\\
 & (0.0007) & (0.0007) & (0.0007)\\
Rain (z) &  & 0.0258*** & 0.2545***\\
 &  & (0.0004) & (0.0100)\\
Mean temperature (10s) &  & -0.0023*** & 0.0654***\\
 &  & (0.0002) & (0.0028)\\
Rain squared &  &  & -0.0410***\\
 &  &  & (0.0004)\\
Temperature squared &  &  & -0.0001***\\
 &  &  & \vphantom{1} (0.0000)\\
Rain times Temperature &  &  & -0.0007***\\
 &  &  & (0.0000)\\
\textbf{Fixed effects:} & \textbf{} & \textbf{} & \textbf{}\\
Village & Yes & Yes & Yes\\
Year & Yes & Yes & Yes\\
\midrule
Observations & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is estimated agricultural yield, defined as tons per hectare. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

To put the size of the coefficient in context, it is worth digging a little deeper into the exposure variable. The mean within-village absolute deviation in wind exposure is approximately 2.42 days, meaning that the average change in agricultural productivity from year-to-year due to nothing but changes in wind patterns carrying particulate matter is around 1.5 percent relative to the mean. This implies that swings in agricultural productivity -- due to wind direction -- of several percent are probably quite common, since the absolute deviation includes deviations below and above the mean.

@@

\caption{\label{tab:yieldtablepm}Particulate matter (PM 2.5) and agricultural productivity}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
\toprule
  & (1) & (2) & (3)\\
\midrule
Particulate matter & -0.042*** & -0.045*** & -0.041***\\
(log PM 2.5) & (0.003) & (0.003) & (0.003)\\
Weather & No & Yes & No\\
Weather (expanded) & No & No & No\\
\textbf{Fixed effects:} & \textbf{} & \textbf{} & \textbf{}\\
Village & Yes & Yes & Yes\\
Year & Yes & Yes & Yes\\
\midrule
Observations & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is estimated agricultural yield, defined as tons per hectare. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

Of course, it is not wind, per se, that we are interested in. Instead, we are actually interested in better understanding the effects of exposure to pollution on agricultural productivity. Table \ref{tab:yieldtablepm} presents the same regressions as Table \ref{tab:yieldtable}, with a key difference: instead of wind exposure, it uses log average concentration of \(\mathrm{PM 2.5}\) -- from Hammer et al. (2020) -- during the season as the key explanatory variable. There is again a consistent, negative relationship between particulate matter and agricultural productivity. Since both variables are logged, the coefficients are similar to elasticities, indicating that a one-percent increase in pollution leads to an approximately 0.04 percent decrease in agricultural productivity.

Recall, however, the results from above: coal plants were placed in places with higher average agricultural productivity. This means that the simple regressions in Table \ref{tab:yieldtablepm} might actually be biased upwards: although pollution might negatively affect productivity, pollution is higher in more productive areas due to the endogenous placement of coal plants in the study areas. One possible way to recover the actual effect of pollution on agricultural productivity is with an instrument. If, conditional on the fixed effects and weather variables, wind direction is exogenous and only affects agricultural productivity through pollution, then it is a valid instrument for pollution.

@@

The second-stage results first indicate that \(\mathrm{PM 2.5}\) is a very strong predictor of agricultural productivity. Using the results in column three, a one percent increase in particulate matter leads to a 0.62 percent decrease in agricultural productivity. To put this in further perspective, the mean particulate matter concentration is just short of 40 (39.68) in the monsoon sample, while the within-village absolute deviation in particulate matter is 3.98, meaning year-to-year swings of several percent in agricultural productivity due to pollution exposure are common.

Column four presents one last specification; it includes a much more flexible specification for weather. Specifically, I turn the linear z-score weather variable into bins for each month of the season. I then interact this rainfall variable with the temperature variable. The coefficient in column four shows almost not change from those in column three, relieving some concerns that weather may not be correctly specified, leading to a spurious correlation between particulate matter and agricultural productivity.

@@

The second robustness check includes leads of pollution and wind direction. I present these results in Table \ref{tab:yieldtableleads}. Neither a one-year nor a two-year lead predicts current agricultural productivity, which supports the identification assumptions.

>>>>

\title{Air pollution and agricultural productivity in a developing country\footnote{I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper, as well as conference attendees at the 2023 KER International Conference. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}}
\author{Joshua D. Merfeld\footnote{KDI School of Public Policy and Management and IZA; \href{mailto:merfeld@kdis.ac.kr}{\nolinkurl{merfeld@kdis.ac.kr}}}}
\date{2023-07-13}

@@

While there may be positive effects from this increase in power generation -- for example, electricity may lead to economic and productivity growth (Dinkelman 2011; Kline and Moretti 2014; Rud 2012; Van de Walle et al. 2017), though electrification alone may not be a sufficient condition (K. Lee, Miguel, and Wolfram 2020) -- these improvements do not come without downsides, especially when it comes to power derived from coal. Pollutants from coal contribute to climate change and the phase out of coal energy is, therefore, a major climate goal (IEA 2022). Emissions from the production and burning of coal are also directly harmful to human health and the environment; pollutants like sulfur dioxide, nitrogen oxides, methane, and mercury are particularly harmful.\footnote{www.eia.gov/energyexplained/coal/coal-and-the-environment.php} While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.\footnote{www.npr.org/2022/08/15/1117560560/a-rising-demand-for-coal-amidst-war-in-ukraine}

@@

In this paper, I study the effects of pollution on agricultural productivity in India. Previous work has suggested that there may be harmful direct effects of air pollution (Heck et al. 1982; Marshall et al. 1997) and water pollution (Reddy and Behera 2006) on agricultural productivity, including work showing that gold mines -- which release large amounts of pollutants into the surrounding environment -- lead to lower agricultural productivity in Ghana (Aragón and Rud 2016). These effects can work through the well-known effects on human health and productivity, but also directly through land productivity. Black carbon in soot, for example, can directly affect solar radiation reaching plants on the surface (Ramanathan and Carmichael 2008; Burney and Ramanathan 2014), while, on the other hand, sulfur in some pollution sources can actually be productivity enhancing (Sanders and Barreca 2022). Relative to previous work in economics, this paper tackles pollution more broadly, before turning to a specific source of pollution: coal plants. I show that air pollution leads to lower agricultural productivity in India and that pollution from coal plants is particularly harmful to agriculture.

@@

The main goal of this paper is to examine whether exposure to pollution affects agricultural productivity, as well as to provide evidence about the effects of pollution from coal plants, in particular. I use several sources of data, which I describe here and list in Table \ref{tab:data} of the appendix.

@@

The second dataset includes agricultural productivity for both the monsoon and winter seasons, from 2002 to 2018. To match other data, I only use the data up to 2013. This data comes from Gangopadhyay et al. (2022) and estimates land productivity (i.e.~yield, in tons per hectare) for the major crops in India, using satellite-derived data. While there is a wealth of literature on measurement error in land (Desiere and Jolliffe 2018; Abay et al. 2022), the authors use a static cropland mask for the productivity estimates. This means that any measurement error will hopefully be constant across time and subsumed by the village fixed effects. While the measure of output is also noisy, the use of satellite-derived data and a static cropland mask reduces concerns about measurement error being responsible for the results below, especially since common sources of measurement error are unlikely to be correlated with wind direction. The authors define the monsoon season as June to October and the winter season as November to March. Due to large differences in the crops grown and the production technologies used across seasons, I focus on only the monsoon season and keep the authors' definition when matching across data below. The data on agricultural productivity is publicly available from Nature's data-sharing website.\footnote{springernature.figshare.com} Since the main regressions below use village fixed effects, any differences in cropping patterns across areas should not bias estimates on changes, as long as these patterns do not change markedly in response to annual pollution levels.

@@

To measure exposure to pollution from point sources around coal plants, I first locate all village centroids located within 30km from a coal plant in a given year. I choose 30km due to previous research on the effects of (air) pollution (Aragón and Rud 2016; Bencko and Symon 1977; Li and Gibson 2014). I then calculate the direction from these point sources to all village centroids within that 30km radius. To define exposure, I pull daily wind direction data from the National Center for Atmospheric Research.\footnote{climatedataguide.ucar.edu/} For each day, I document whether the wind is blowing towards each village centroid\footnote{I define ``towards'' as within five degrees to help take into account that I am using village centroids. The wind data includes both the x and y components of wind speed, which I can use to calculate angles from point sources to village centroids.} and then temporally aggregate this daily data depending on the temporal definition of the corresponding outcome. For example, agricultural productivity is defined across five months (i.e.~the monsoon season is from June to October) so I count the total days a given village is exposed to wind blowing from \emph{any} coal plant within 30km during those five months.

The top panel in Figure \ref{fig:windexample} in the appendix is an example wind direction raster. The raster shows wind direction for the entirety of India on January 1st, 2010, as well as the location of coal power plants on that date. The prevailing winds on the date differ across the country and, though not shown in the figure, across days. This means that the overall exposure for a given area to coal plant emissions changes across time.

For pollution data, I use estimated particulate matter (PM). Specifically, I use estimates of PM 2.5, which includes particles with a diameter of 2.5 micrometers or less. This data comes from Hammer et al. (2020) and is available as a raster with a resolution of 0.01 degrees. I aggregate this data to the village level by extracting the mean PM for each feature in the shapefile. I do this separately for each month from June 2002 to October 2013. The bottom panel in Figure \ref{fig:windexample} in the appendix is an example of the PM data. The figure shows the estimated PM 2.5 from for each village in Guna district of Madhya Pradesh on January 1st, 2010.

While I use PM 2.5 as the estimated level of pollution, it is important to clarify that I cannot explicitly say that PM 2.5, per se, is responsible for any effects. Pollution includes a variety of different components, including PM 2.5, PM 10, sulfur dioxide, and nitrogen dioxide, to name but a few. While I use PM 2.5 as a proxy for pollution, it is possible that other components are responsible for the effects. Additionally, the PM 2.5 data is estimated using satellite data and a machine learning algorithm. While this is a common approach in the literature ( see Hammer et al. (2020) for more details), it is important to note that the estimates are not direct measurements of pollution. However, a valid instrumental variable can help for any measurement error present in the data (Glewwe 2012; Gillen, Snowberg, and Yariv 2019).

To help understand some of the mechanisms driving the relationship between exposure to emissions and agricultural productivity, I also look at the differential effects by the labor intensity of the predominant crop in each area. For this, I use data from Monfreda, Ramankutty, and Foley (2008), who estimate area harvested for different crops across the global around the year 2000. I identify areas where rice (paddy) is the most commonly grown crop, under the assumption that rice is more labor intensive than other crops (Vollrath 2011; Michler 2020). Michler (2020), for example, shows that rice, on avearge, uses almost 1,800 hours per hectare, while the second-highest crop in his data, cotton, juses just 850. He also estimates that wheat, another common crop in India, uses just 600 hours per hectare.

@@

\caption{\label{tab:plantresultstable}Local characteristics and the construction of coal plants}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3.5cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{1991 census} & \multicolumn{2}{c}{2001 census} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
  & 1991 & 2001 & 2001 & 2011\\
\midrule
Population (log) & 0.012*** & 0.009*** & 0.015*** & 0.006***\\
 & (0.001) & (0.000) & (0.000) & \vphantom{1} (0.000)\\
Area (log) & -0.010*** & -0.005*** & -0.010*** & -0.005***\\
 & (0.001) & (0.000) & (0.000) & (0.000)\\
Urban pop (prop) & -0.029 & 0.006 & -0.003 & -0.007\\
 & (0.037) & (0.035) & (0.022) & (0.018)\\
Ag. productivity &  &  & 0.014*** & 0.028***\\
 &  &  & (0.001) & (0.001)\\
Pollution (log PM 2.5) &  &  & 0.051*** & 0.016***\\
 &  &  & (0.002) & (0.001)\\
Sub-sample & all & no plant & all & no plant\\
\midrule
Observations & 283,021 & 264,265 & 508,786 & 469,005\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: State fixed effects are included in regressions. Robust standard errors are in parentheses. The outcome in the first column is whether the village is within 30km of a coal plant in 1991. The second column is whether a village in 1991 will have a coal plant in 2001, conditional on not having one in 1991. The last two columns are similarly defined, except using 2001 and 2011 as the years. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

The inclusion of rainfall and temperature is meant to reduce concerns that wind direction may not affect agricultural productivity but, instead, be correlated with it through changes in weather. For example, prevailing winds from one direction may be associated with dry weather, while winds from another direction may be more likely to carry moisture and rain. It is not a priori clear how this would bias the estimates with respect to pollution and coal plants, however. The differences in weather patterns is very context specific and can differ even for villages located near one another.

I believe there are two reasons this is unlikely to be the case here. First, villages located near one another have very similar weather patterns, but wind direction leads to large differences in pollution exposure even for villages right next to one another. Second, in some specifications, I include very flexible specifications for weather. These inclusions do not lead to very large changes in the estimated relationship between pollution and agricultural productivity, suggesting that weather is unlikely to be responsible for a spurious correlation between productivity and pollution in the main results.

It is worth taking the time to validate the use of wind direction as a measure of exposure to pollution. Consider the data used for the regressions presented in Table \ref{tab:yield3ivmainfirststage} of the appendix. The outcome variable is particulate matter -- specifically, \(\mathrm{PM 2.5}\), which is particulate matter no larger than 2.5 micrometers in diameter -- which comes from Hammer et al. (2020). \(\mathrm{PM 2.5}\) is one of the harmful byproducts from many industrial processes, including power generation. Particulate matter is a major component of the pollution emitted by coal plants, along with sulfur dioxide, different types of nitrogen oxides, and mercury.\footnote{www.epa.gov/airmarkets/power-plants-and-neighboring-communities} This data is available at the monthly level, but I aggregate variables to the season variable since that is the level of analysis in the main results.

Both the particulate matter data and the wind data is at the village level, which means the regressions in Table \ref{tab:yield3ivmainfirststage} are at the village level. As such, the geographic fixed effects are village fixed effects and the temporal fixed effects are month fixed effects. Since the exposure variable is at the village level and we follow villages over time, the standard errors are also clustered at the village level. I add more and more flexible specifications for weather, with the most flexible specification in the fourth column. Note that the relationship between wind direction and particulate matter is very similar across all four columns, suggesting that the inclusion of weather does not change the relationship between wind direction and pollution. This is important because it suggests that the use of wind direction as a proxy for pollution exposure is not biased by weather patterns.

@@

A key question is how pollution affects agricultural productivity. Specifically, does pollution directly affect labor productivity, land productivity, or both? Unfortunately, there is a lack of data that would allow me to directly test these mechanisms, since I do not have data on the amount of labor or land used in each village. Instead, I look at heterogeneity in the effects of pollution based on the predominant crop grown in each village.

Michler (2020) presents estimates of labor intensity for several major crops in India. The most labor-intensive crop is rice, which uses almost 1,800 hours per hectare. The second-most labor-intensive crop is cotton, which uses just 850 hours per hectare. Wheat, another common crop in India, uses just 600 hours per hectare. This suggests that pollution should be more harmful for rice than for other crops if there are direct impacts on labor productivity.

To identify this possible mechanism, I identify areas that grow predominantly rice. Specifically, I estimate the following regression:
\begin{gather} 
y_{it} = \alpha_{i} + \gamma_{t} + \omega PM_{it} + \phi rice_{it} + \beta PM_{it}\times rice_{it} + \delta rain_{it} + \psi temp_{it} + \varepsilon_{it},
\end{gather}
where \(PM_{it}\) denotes the effect of pollution in non-rice areas and \(PM_{it}\times rice_{it}\) is the differential effect in rice-growing areas. Here, I create an indicator variable for rice areas that is time invariant, so \(rice_{it}\) drops out of the estimation. I instrument for \(PM_{it}\) and for \(PM_{it}\times rice_{it}\) using wind and wind interacted with the indicator for rice.

While this is a useful first step, it is not a perfect identification strategy. First, there could be other differences in the types of areas that grow rice, and these differences could actually be responsible for any heterogeneity. Second, this strategy does not allow me to say anything about the effects of pollution on land productivity.

@@

I first present the reduced form results for agricultural productivity. The outcome for all regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. The first set of results are in Table \ref{tab:yieldtable}. The unit of analysis is the village-season -- from 2002 to 2013 -- focusing only on the monsoon (kharif) season. The data used to construct the yield measure is calculated using five months of data, so the wind variable is defined for the same five months and is the sum of days in which the wind is pointed towards the village centroid from a coal plant.

The first column presents the most simple results, with only the wind exposure variable and the fixed effects. The second column adds a rainfall and average temperature variable. While I leave them out of the table for simplicity, the coefficients are in the expected direction: rainfall is positive and average temperature is negative. The third column adds higher-order weather variables for both rainfall and temperature, as well as the interaction between them. This attenuates the relationship between wind and productivity somewhat, but the coefficient remains negative and statistically significant.

The final column includes the most flexible form for the weather variables. The z-score for rainfall is turned into bins -- with cutoffs at \(z=-1\) and \(z=1\) -- based on rainfall in each month of the season. In other words, over the five months, I create 15 different indicator variables for rainfall based on the z-score in each month. One of these indicators drops out for each month of the year, leaving 10. I also include interactions between all of these and temperature, as well as between the bins and temperature squared. The coefficient on wind remains negative and statistically significant, and is actually slightly larger in magnitude than the results in column three, which is a much less flexible specification for weather. This leads me to conclude that the results are not being driven by weather patterns.

@@

\caption{\label{tab:yieldtable}Wind direction and agricultural productivity}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}c}
\toprule
  & (1) & (2) & (3) & (4)\\
\midrule
wind days & -0.0078*** & -0.0065*** & -0.0051*** & -0.0053***\\
 & (0.0007) & (0.0007) & (0.0007) & (0.0007)\\
weather & No & Yes & Yes & No\\
weather (expanded) & No & No & Yes & No\\
weather (expanded, bins) & No & No & No & Yes\\
\textbf{fixed effects:} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
village & Yes & Yes & Yes & Yes\\
year & Yes & Yes & Yes & Yes\\
\midrule
observations & 1,266,922 & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is estimated agricultural yield, defined as tons per hectare. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

To put the size of the coefficient in context, it is worth digging a little deeper into the exposure variable. The mean within-village absolute deviation in wind exposure is approximately 2.42 days, meaning that the average change in agricultural productivity from year-to-year due to nothing but changes in wind patterns carrying particulate matter is around 1.5 percent relative to the mean. This implies that swings in agricultural productivity -- due to wind direction -- of several percent are probably quite common, since the absolute deviation includes deviations above and below the mean.

@@

\caption{\label{tab:yieldtablepm}Particulate matter (PM 2.5) and agricultural productivity}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{5cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}c}
\toprule
  & (1) & (2) & (3) & (4)\\
\midrule
Particulate matter (log PM 2.5) & -0.042*** & -0.045*** & -0.008** & -0.041***\\
 & (0.003) & (0.003) & (0.003) & (0.003)\\
Weather & No & Yes & Yes & No\\
Weather (expanded) & No & No & Yes & No\\
weather (expanded, bins) & No & No & No & Yes\\
\textbf{Fixed effects:} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
Village & Yes & Yes & Yes & Yes\\
Year & Yes & Yes & Yes & Yes\\
\midrule
Observations & 1,266,922 & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is estimated agricultural yield, defined as tons per hectare. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

Of course, it is not wind, per se, that we are interested in. Instead, we are actually interested in better understanding the effects of exposure to pollution on agricultural productivity. Table \ref{tab:yieldtablepm} presents the same regressions as Table \ref{tab:yieldtable}, with a key difference: instead of wind exposure, it uses log average concentration of \(\mathrm{PM 2.5}\) -- from Hammer et al. (2020) -- during the season as the key explanatory variable. There is again a consistent, negative relationship between particulate matter and agricultural productivity. Since both variables are logged, the coefficients are similar to elasticities, indicating that a one-percent increase in pollution leads to an approximately 0.04 percent decrease in agricultural productivity in three of the columns.

Recall, however, the results from above: coal plants were placed in places with higher average agricultural productivity. This means that the simple regressions in Table \ref{tab:yieldtablepm} might actually be biased upwards: although pollution might negatively affect productivity, pollution is higher in more productive areas due to the endogenous placement of pollution sources in the study areas. One possible way to recover the actual effect of pollution on agricultural productivity is with an instrument. If, conditional on the fixed effects and weather variables, wind direction is exogenous and only affects agricultural productivity through pollution, then it is a valid instrument for pollution.

@@

The second-stage results first indicate that \(\mathrm{PM 2.5}\) is a very strong predictor of agricultural productivity. Using the results in column three, a one percent increase in particulate matter leads to a 0.62 percent decrease in agricultural productivity. To put this in further perspective, the mean particulate matter concentration is just short of 40 (39.68) in the monsoon sample, while the within-village absolute deviation in particulate matter is 3.98, meaning year-to-year swings of several percent in agricultural productivity due to pollution exposure alone are common.

Column four presents one last specification; it includes a much more flexible specification for weather. This is the same binned specificaiton as I describe above. The coefficient in column four shows almost not change from those in column three, again relieving some concerns that weather may not be correctly specified, leading to a spurious correlation between particulate matter and agricultural productivity.

@@

The second robustness check includes leads of pollution and wind direction. I instrument for future pollution using future wind direction, which mirrors the strategy employed in the main results. I present these results in Table \ref{tab:yieldtableleads}. Neither a one-year nor a two-year lead predicts current agricultural productivity, which supports the identification assumptions.
